Trail Saving on Backtrack

### 2020-02-21 paper.tex
added abstract

\documentclass[runningheads]{llncs}
\usepackage{times}
\usepackage{amsmath}
\usepackage{amssymb}
\usepackage[noend]{algcompatible}
\usepackage[ruled,linesnumbered,noend,oldcommands]{algorithm2e}
\usepackage[T1]{fontenc}
\usepackage{hyperref}
\usepackage{xspace}
\usepackage{graphicx}
\usepackage{listings}
\usepackage{xcolor}
\lstset { %
  language=C++,
  % backgroundcolor=\color{black!5}, % set backgroundcolor
  basicstyle=\footnotesize,% basic font setting
}

\renewcommand{\note}[1]{{\color{red}{#1}}}

@@

\begin{document}
\maketitle
\begin{abstract}
\end{abstract}

@@

\note{Fix later---also \cite{DBLP:journals/jsat/TakRH11} can't be
  first use of trail savings?}

@@

The vast majority of modern SAT solvers that are used to solve
real-world problems are based on the conflict-driven clause learning
(CDCL) algorithm. In a CDCL SAT solver, backtracking occurs after
every conflict, where all literals from one or more decision levels
become unassigned, before the solver resumes making decisions and
performing unit propagations. Traditionally, CDCL solvers would
backtrack non-chronologically to the conflict level, which is the
second highest decision level remaining in the conflict clause after
conflict analysis has resolved away all but one literal from the
current decision level \cite{DBLP:conf/dac/MoskewiczMZZM01}. Recently,
however, it has been shown that backtracking chronologically
(i.e. backtracking one level only after conflict analysis)
\cite{DBLP:conf/lpar/2013,DBLP:conf/sat/NadelR18} can be effective on
many instances and was implemented in solvers that won the last two
SAT competitions. Although chronological backtracking breaks some of
the conventional invariants of CDCL solvers, it has been formalized
and proven correct \cite{DBLP:conf/sat/MohleB19}.

There are at least two possible side-effects of chronological
backtracking which may explain its effectiveness. It has been observed
that when a solver backtracks non-chronologically after a conflict,
many of the literals that are unassigned during the backtrack will be
re-assigned again in roughly the same order when the solver
continues. By using chronological backtracking, the solver keeps a lot
more of its partial assignment intact and saves from having to
reconstruct a lot of the trail via propagation that it otherwise would
have. Saving from having to reconstruct portions of the trail after
backtracking was first introduced in the context of restarts
\cite{DBLP:journals/jsat/TakRH11}. The second side-effect of
chronological backtracking is that it forces the solver to stay in the
same local neighbourhood during search, since it unassigns less of the
partial assignment and one could argue that this might lead to finding
additional conflicts more quickly or reaching a satisfying assignment
(if one exists) more quickly. However, this second side-effect also
leads to less search flexibility, since the solver does not jump out
of local neighbourhoods as often and does not obey the variable
selection heuristic. We will argue that it is the first side-effect
that leads to the effectiveness of chronological backtracking and not
the second by developing a technique which also saves from having to
repeat propagations, but allows the search to remain flexible and obey
the variable selection heuristic.

@@

\section{Background}
\note{elaborate backtrack after learning a clause. Also describe
  properties of the trail (a) reason valid every reason clause is made
  unit by its prefix, (b) non-contradictory, (c) non-conflicting (no
  clause is falsified), and (d) propagation complete }

We will assume the reader is familiar with SAT solving and the basics
of the CDCL algorithm \cite{DBLP:series/faia/SilvaLM09}. We will use
the notion of a trail to represent a sequence of literals
corresponding to the current partial assignment. The literals appear
on the trail in the order in which they were assigned. Associated with
every literal on the trail is a decision level. Decision level 0
represents all literals which are implied by the formula itself and
subsequent decisions start the beginning of a new decision level. When
a literal is made true by unit propagation, the literal is given its
decision level and a reason clause, where the reason clause is the
clause that was made unit by propagation in order to imply this
literal. Most modern solvers use the two-literal watch scheme to make
unit propagation more efficient, where each clause is watched by two
non-false literals. When a watcher for a clause becomes false, a
replacement is searched for in the clause. If no such replacement
exists and the other watcher is true or unassigned, then the clause is
unit. If no such replacement exists and the other watcher is also
false, then the clause is a conflict and needs to be analyzed to
produce a learnt clause.

\section{Trail Saving}
\note{
  \begin{enumerate}
  \item Use simple backtrack and save to motivate and state the
      invariant. 
  \item Introduce notation for concatenating two trails.
  \item Name and state the invariant in text (e.g. a
      defn)---concenating the cached trail with the current trail
      yields a trail that is always reason valid, but not necessarily
      non-contradictory, non-conflicting, nor propagation complete).
  \item Proof (simple) that the invariant holds for caching on
      backtrack.
  \item On backtrack at least the asserted literal and any new
      propagants at the backtrack level are added.
  \item Prove that the invariant holds (???do we need an invariant or
      simply a property of the augmented trail) for trails with extras
      inbeween.
      $T_{\mathit{old}} + T_{\mathit{newlyAdded}} +
      T_{\mathit{saved}}$ also satisfies the invariant
  \item re-acheving propagation completeness for the augmented trail
      (a) only add one level of the cached trail at a time, and do
      unit prop after (b) only add non-contradicted decision from the
      old trail. Also note that conflicts are valid and can be learnt
      from even if the augmented trail is not propagation complete. 
  \item Finally figure out a way to formalize the extra technique of
      storing on top after repeated backtracking. 
  \item Explain how freedom for the search is maintained by makeing
      decisions then consulting the trail.
  \item 
  \end{enumerate}
}

@@

\begin{algorithm}
  	\TitleOfAlgo{propagate()}
	\begin{algorithmic}[1]
        \WHILE{queue not empty}
        \STATE { }
        \STATE restoreTrailCache()
        \STATE { }
        \STATE $l := $ next literal on queue
        \FOR{each clause $c \in $ watchlist($\neg l$)}
		\FOR{each literal $x \in c$}
        \IF{$val($x$) \neq$ false and x is not already a watcher of c}
        \STATE let $x$ be a watcher of $c$ and add $c$ to watchlist($\neg x$)
        \STATE break
        \ENDIF
		\ENDFOR
		\IF{no new watcher was found}
        \IF{other watcher $x_2$ of $c = $ false}
        \STATE return c as conflict
        \ELSIF{other watcher $x_2$ of $c$ is unassigned}
        \STATE enqueue $x_2$
        \ENDIF
		\ENDIF
        \ENDFOR
        \ENDWHILE
	\end{algorithmic}
\end{algorithm}

\begin{algorithm}
  	\TitleOfAlgo{restoreTrailCache()}
	\begin{algorithmic}[1]
        \IF{trail saving turned on}
		\STATE{let tc $=$ top of trail cache}
		\IF{val(top of trail cache) $=$ true}
        \WHILE{trail cache not empty}
        \STATE{move top of trail cache (tc) to next literal}
        \IF{reason(tc) is null}
        \STATE{break}
        \ELSIF{val(tc) $=$ false}
        \STATE{return conflict with reason(tc)}
        \ELSIF{val(tc) $=$ unassigned}
        \STATE{enqueue tc with reason reason(tc)}
        \ENDIF
        \ENDWHILE
		\ENDIF
        \ENDIF
	\end{algorithmic}
\end{algorithm}

>>>>

\documentclass[runningheads]{llncs}
\usepackage{times}
\usepackage{amsmath}
\usepackage{amssymb}
\usepackage{algorithm}
\usepackage[noend]{algpseudocode}
%\usepackage[noend]{algcompatible}
%\usepackage[ruled,linesnumbered,noend,oldcommands]{algorithm2e}
\usepackage[T1]{fontenc}
\usepackage{hyperref}
\usepackage{xspace}
\usepackage{graphicx}
\usepackage{listings}
\usepackage{xcolor}
\usepackage{theorem}
\algrenewcommand\algorithmicprocedure{}
\algrenewcommand\algorithmicthen{}
\algrenewcommand\algorithmicdo{}

\lstset { %
  language=C++,
  % backgroundcolor=\color{black!5}, % set backgroundcolor
  basicstyle=\footnotesize,% basic font setting
}

\renewcommand{\note}[1]{{\color{red}{#1}}}
\newcommand{\fs}[1]{\fontsize{#1}{#1}\selectfont}
\newcommand{\fss}[2]{\fontsize{#1}{#2}\selectfont}
\newcommand{\sat}{SAT\xspace}
\newcommand{\trail}{\ensuremath{\mathcal{T}}}
\newcommand{\trailIdxT}[2]{\ensuremath{\iota_{#1}(#2)}}
\newcommand{\trailIdx}[1]{\ensuremath{\iota(#1)}}
\newcommand{\range}[2]{#1\ldots#2}
\newcommand{\dlevel}[1]{\ensuremath{\mathit{decLvl}(#1)}}
\newcommand{\dlevels}{\ensuremath{\mathit{decLvls}}}
\newcommand{\var}{\text{var}}
\newcommand{\true}{\textsc{true}\xspace}
\newcommand{\false}{\textsc{false}\xspace}
\newcommand{\reason}[1]{\ensuremath{\mathit{reason}(#1)}}
\newcommand{\reasonsave}[1]{\ensuremath{\mathit{reason_{\mathit{save}}(#1)}}}
\newcommand{\litsave}{\ensuremath{\mathit{l_{\mathit{save}}}}}
\newcommand{\formula}{\ensuremath{\mathcal{F}}}
\renewcommand{\implies}{\rightarrow}
\newcommand{\uipcls}{C_{\textit{1-UIP}}}
\newcommand{\deepestLvl}{L_{\textit{deep}}}
\newcommand{\deepestLit}{\ell_{\textit{deep}}}
\newcommand{\btL}{L_{\textit{back}}}
\newcommand{\cbt}{C-bt\xspace}
\newcommand{\trailsave}{\trail_{\mathit{save}}}

\newtheorem{Obs}{Observation}
\newtheorem{Cor}{Corollary}
\newtheorem{defn}{Definition}
\newtheorem{thm}{Theorem}
\newtheorem{ex}{Example}
\newcommand{\whitebox}{\raisebox{.5ex}{\fbox{\hspace*{.2ex}}}}

@@

\begin{document}
\maketitle
\begin{abstract}
    A CDCL \sat solver can backtrack a large distance when it learns a
    new clause, e.g, when the new learnt clause is a unit clause the
    solver has to backtrack to level zero. When the length of the
    backtrack is large the solver can end up reproducing many of the
    same decisions and propagations when it redescent the search
    tree. Motivated by this potential redundancy, partial backtracking
    (chronological backtracking) has been proposed. This technique
    allows the solver to reduce the length of its backtrack, e.g.,
    allowing the solver to backtrack to the previous level after a
    conflict is found. However, this technique has its shortcomings
    and only tends to work when limits are placed on its
    application. In this paper we present a new trail saving technique
    that makes a copy of the part of the trail that was backtracked
    over. This saved copy can then be used to improve the efficiency
    of the solver's subsequent redescent. Although our technique does
    not save as much work as chronological backtracking it also does
    not have chronological backtracking's shortcomings. Furthermore,
    the saved trail also provides the solver the ability to look ahead
    along the previous trail. This can be exploited to improve its
    efficiency. Our empirical results show that our technique often
    yields superior performance over chronological backtracking, and
    that is is able to improve the performance of state of the art
    solvers.

\end{abstract}

@@

@@

The vast majority of modern SAT solvers that are used to solve
real-world problems are based on the conflict-driven clause learning
(CDCL) algorithm. In a CDCL SAT solver, backtracking occurs after
every conflict, where all literals from one or more decision levels
become unassigned, before the solver resumes making decisions and
performing unit propagations. Traditionally, CDCL solvers would
backtrack non-chronologically to the conflict level, which is the
second highest decision level remaining in the conflict clause after
conflict analysis has resolved away all but one literal from the
current decision level \cite{DBLP:conf/dac/MoskewiczMZZM01}. Recently,
however, it has been shown that backtracking chronologically
\cite{DBLP:conf/lpar/JiangZ13} (i.e. backtracking one level only after
conflict analysis)
\cite{DBLP:conf/sat/NadelR18,DBLP:conf/sat/MohleB19} can be effective
on many instances and was implemented in solvers that won the last two
SAT competitions. Although chronological backtracking breaks some of
the conventional invariants of CDCL solvers, it has been formalized
and proven correct \cite{DBLP:conf/sat/MohleB19}.

There are at least two possible side-effects of chronological
backtracking which may explain its effectiveness. It has been observed
that when a solver backtracks non-chronologically after a conflict,
many of the literals that are unassigned during the backtrack will be
re-assigned again in roughly the same order when the solver
continues. By using chronological backtracking, the solver keeps a lot
more of its partial assignment intact and saves from having to
reconstruct a lot of the trail via propagation that it otherwise would
have. Saving from having to reconstruct portions of the trail after
backtracking was first introduced in the context of restarts by van
der Tak et al. \cite{DBLP:journals/jsat/TakRH112}. The second
side-effect of chronological backtracking is that it forces the solver
to stay in the same local neighbourhood during search, since it
unassigns less of the partial assignment and one could argue that this
might lead to finding additional conflicts more quickly or reaching a
satisfying assignment (if one exists) more quickly. However, this
second side-effect also leads to less search flexibility, since the
solver does not jump out of local neighbourhoods as often and does not
obey the variable selection heuristic. We will argue that it is the
first side-effect that leads to the effectiveness of chronological
backtracking and not the second by developing a technique which also
saves from having to repeat propagations, but allows the search to
remain flexible and obey the variable selection heuristic.

@@

\section{Background}
A propositional formula $\formula$ expressed in Conjunctive Normal
Form (CNF), contains a set of variables $V$. A literal is a variable
$v\in V$ or its negation $\lnot v$, and for a literal $l$ we let
$\var(l)$ denote its underlying variable. A CNF consists of a
conjunction of clauses, each of which is a disjunction of literals. We
often view a clause as being a set of literals and employ set
notation, e.g., $\ell\in C$ and $C'\subset C$.

We assume the reader is familiar with the basic operations of CDCL
\sat solvers. A good source for this background is
\cite{DBLP:series/faia/SilvaLM09}.

\paragraph{Trails.}
CDCL \sat solvers maintain a trail which is the sequence of literals
that have currently been assigned \true by the solver. During its
operation a \sat solver will add newly assigned literals to the end of
the trail, and on backtrack remove literals from the end of the trail
(unassigning them as they are removed).

A \sat solver's trail satisfies a number of conditions that we will
discuss below. However, in we will need some additional flexibility in
our definitions, as we will be sometimes be working with trails that
would never be constructed by a \sat solver. Hence, we define a
\emph{trail} to be a sequence of literals each of which is either a
\emph{decision} literal or an \emph{implied} literal, and each of
which has a \emph{reason}. These two types of literals are
distinguished by their \emph{reasons}. Decision literals $d$ have a
null reason, $\reason{d} = \varnothing$. Implied literals $l$ have as
a reason a clause of the formula $\formula$,
$\reason{l} = C\in\formula$. ($\reason{l}$ can be a learnt clause that
has been added to $\formula$).

If literal $\ell$ is on the trail $\trail$ let
$\trailIdxT{\trail}{\ell}$ denote its index on the trail, i.e,
$\trail[\trailIdxT{\trail}{\ell}] = \ell$. For convenience, we also let
$\trailIdxT{\trail}{\ell} = \trailIdxT{\trail}{\lnot \ell} =
\trailIdxT{\trail}{\var(\ell)}$ even though neither $\lnot \ell$ nor
$\var(\ell)$ are actually on $\trail$. If $x$ and $y$ are both on the
trail and $\trailIdxT{\trail}{x} < \trailIdxT{\trail}{y}$ we say that
\textit{$x$ appears before $y$ on the trail}. For convenience, when
the trail being discussed in clear from context we simply write
$\trailIdx{}$ instead of $\trailIdxT{\trail}{}$.

Each literal $\ell\in\trail$ has a decision level $\dlevel{\ell}$
which is equal to the number of decision literals appearing before
$\ell$ on the trail plus one; hence, $\dlevel{d}=1$ for the first
decision literal $d\in\trail$. The set of literals on $\trail$ that
have the same decision level forms a \emph{contiguous}
subsequence\footnote{Our approach uses standard trails in which the
  decision levels are contiguous. Chronological backtracking
  \cite{DBLP:conf/lpar/JiangZ13,DBLP:conf/sat/NadelR18,DBLP:conf/sat/MohleB19}
  generates trails with non-contigous decision levels} that starts
with a decision literal $d_i$ and ends just before the next decision
literal $d_{i+1}$. We will often need to refer to different decision
level subsequences of $\trail$. Hence, we let $\trail[[i]]$ denote the
subsequence of literals at decision level $i$; and let
$\trail[[\range{i}{j}]]$ denote the subsequence of literals at
decision levels $k$ for $i\leq k\leq j$. Now we define the
following properties that a trail $\trail$ can have.
\begin{description}
\item[non-contradictory:] A variable cannot appear in both polarities
    in the trail: $l\in \trail\implies \lnot l \not \in \trail$.
\item[non-redundant:] A literal can only appear once on $\trail$.
\item[reason-sound:] For each implied literal $l\in\trail$ we have
    that it reason clause has \emph{been made unit by $\trail$
      implying $l$}. This means that the reason clause contains $l$
    and the negation all of its other literals are on $\trail$ before
    $l$:
    $\forall l\in\trail.\, \reason{l}\neq \varnothing \implies
    l\in\reason{l} \land \bigl(\forall x\in\reason{l}.\, x\neq l
    \implies \lnot x \in \trail \land \trailIdx{\lnot x} <
    \trailIdx{l}\bigr)$.
\item[propagation-complete:] Unit propgation has been run to
    completion at all decisions levels of $\trail$. This means that
    literals appear on $\trail$ at the first decision level they were
    unit implied. Formally, this can be captured by the condition:
    $\forall i \in \{\dlevel{l}\,|\,l\in \trail\}. \bigl(\exists
    C\in\formula. \mbox{$C$ is made unit by $\trail[[\range{0}{i}]]$
      implying $l$}\bigr) \implies l\in \trail[[\range{0}{i}]]$. With
    non-redundancy $l$ will appear only at the first decision level it
    was implied, without non-redundancy, $l$ might appear more than
    once. Note that this also implies that $\reason{l}$ must contain
    at least one other literal $y\neq l$ with
    $\dlevel{y} = \dlevel{l}$.
\item[conflict free:] No clause of $F$ is falsified by
    $\trail$. Clauses $C\in F$ falsified by $\trail$ are typically
    called \emph{conflicts}.
\end{description}

In CDCL solvers using standard conflict directed backtracking all
properties hold of the prefix of the solver's trail consisting of all
decisions levels but the deepest. The full trail might, however,
contain a conflict at its deepest level so is not necessarily conflict
free. The full trail might also not be propagation-complete, as unit
propagation at the deepest level is typically terminated early if a
conflict is found. It can further be noted that the first four
properties imply that if a clause $C$ is falsified at decision level
$k$, then $C$ must contain at least two literals at level $k$
(otherwise $C$ would have become unit at a prior level and then
satisfied by making its last unfalsified literal \true).

\paragraph{Standard Backtracking.}
In CDCL \sat solving the solver extends its trail by adding new
decision literals followed by finding and adding all unit implied
literals arising from that new decision. This continues until it
reaches a decision level $\deepestLvl$ where a conflict $C$ is found.

In standard backtracking, the solver then constructs a new 1-UIP
clause by resolving away all but one literal at level $\deepestLvl$
from the conflict $C$ using the reason clauses of these literals. (As
noted above $C$ must contain at least two literals at level
$\deepestLvl$). Hence, the new clause $\uipcls$ will contain one
literal $\deepestLit$ at level $\deepestLvl$ and have all of its other
literals a levels less than $\deepestLvl$. The solver then backtracks
to $\btL$ the second deepest level in $\uipcls$. This involves
changing $\trail$ to its prefix $\trail[[\range{0}{\btL}]]$
(unassigning all literals removed from $\trail$). The new clause
$\uipcls$ is made unit by $\trail[[\range{0}{\btL}]]$ implying
$\deepestLit$, so the solver then adds $\deepestLit$ to the trail and
executes another round of unit propagation at level $\btL$, after
which it continues by once again growing the trail with new decisions
and unit implied literals until a new conflict or a satisfying
assignment is found.

In standard backtracking, the difference between the backtrack level,
$\btL$ and the current deepest level $\deepestLvl$ can be very
large. During its new descent from $\btL$ the solver can reproduce a
large number of the same decisions and unit propagations, essentially
wasting work. This potential inefficiency has been noted in prior work
\cite{DBLP:journals/jsat/TakRH11,DBLP:conf/lpar/JiangZ13,DBLP:conf/sat/NadelR18,DBLP:conf/sat/MohleB19}.

In \cite{DBLP:journals/jsat/TakRH11} a technique for reducing the
length of the backtrack during restarts was presented. In restarts,
the solver backtracks to level $0$, and this technique involves
computing a new deeper backtrack level $M> 0$ for which it is known
that on redescent the first $M+1$ levels of the trail will be
unchanged (except perhaps for the ordering of the literals). This
technique removes the redundant work of reproducing the first $M$
trail levels. Unfortunately, with backtracking the trail will be
changed at level $\btL$ ($\deepestLit$ will be newly inserted at this
level). Thus there seems to be no choice but to backtrack to
$\btL$. In this paper we will show that although we have to backtrack
to $\btL$ we can make the subsequent redescent much more efficient.

\paragraph{Chronological Backtracking.}
Chronological bactracking (\cbt) is a technique originally propsed by
Jiang and Zhang \cite{DBLP:conf/lpar/JiangZ13} under the name
\textit{partial backtracking}. Its aim is to reduce the redundant work
that might be done by the \sat solver on its redescent from the
backtrack level $\btL$. The original paper proposed a technique for
backtracking to any level $j$ in the range
$\btL \leq j \leq \deepestLvl{-}1$ (where $\deepestLvl$ is the level
the conflict was discovered. Nadel and Ryvchin
\cite{DBLP:conf/sat/NadelR18} proposed to always backtrack to
$\deepestLvl{-}1$ while M{\"{o}}hle and Biere
\cite{DBLP:conf/sat/MohleB19} returned to the original proposal of
flexibly backtracking any level in the allowed range. Note that the
new 1-UIP clause $\uipcls$ is made unit at every level in this
range. So after backtracking to level $j$ the newly implied literal
$\deepestLit$ is added to the trail with
$\reason{\deepestLit}=\uipcls$, and $\dlevel{\deepestLit}$ is set to
$\btL$ (the second deepest level in $\uipcls$.

This means that the decision levels on the trail are no longer
contiguous, as $\deepestLit$ has a different level than the other
literals at level $j$ (if $j\neq \btL$). This change has a number of
consequences on the \sat solvers operation, all of which were
described in the original paper \cite{DBLP:conf/lpar/JiangZ13}. In
\cite{DBLP:conf/sat/NadelR18} some more practical implementation
methods were developed to overcome this issues, and in
\cite{DBLP:conf/sat/MohleB19} a formal framework for \cbt was
developed under which the changes required to the \sat solver could be
shown to be successful. We refer the reader to the cited papers for
more details.

\section{Potential Drawbacks of Chronological Backtracking}
In this paper we present a new technique that allows the \sat solver
to use standard backtracking, but also allows saving some redundant
work on its redescent.  Our method has more overhead than \cbt so the
first question that must be addressed is why not just use
chronological backtracking.

The reduction in redundant work achieved by \cbt does not come for
free. In particular, some of the work avoided by \cbt is not actually
redundant work; it is work that can actually help the solver. The
obvious evidence for this statement is that in both
\cite{DBLP:conf/sat/NadelR18} and \cite{DBLP:conf/sat/MohleB19} it was
found that it was not optimal to always apply \cbt. In
\cite{DBLP:conf/sat/NadelR18} \cbt was applied only when the length of
the standard backtrack, $\btL-\deepestLvl$ was greater than a
given threshold $T$. In their experiments they found that $T=100$ was
the best value, i.e., \cbt is done only on longer backtracks. In
practice, this meant that \cbt was \emph{hardly ever done}; in our
measurements with their solver only about 3\% of the solver backtracks
were \cbt backtracks. In \cite{DBLP:conf/sat/MohleB19} the value
$T=100$ was also applied. However, they introduced an additional
technique to add some applications of \cbt when the length of the
backtrack is less than $T$. This allowed to utilize \cbt in \%15 of
the backtracks. Nevertheless, the fact \cbt has to be used in a
controlled fashion, indicates that can be benefits in the \sat solver
employing standard backtracking; it cannot be the case that every
redescent is only doing redundant work.

What is this non-redundant work? With standard backtracking the
literal $\deepestLit$ is placed on the trail at the end of $\btL$ and
then unit propagated. This could impact the trail in at least the
following ways. First, some literals might become unit at earlier
levels. This could include decision literals becoming forced which
might compress some decision levels together. Second, different
decisions might be made due to changes in the variable scores arising
from the newly learnt clause. And third, literals might be unit
implied with different reasons. These changes could be sometimes be
major, and they could have a large impact on future
clauses. Unfortunately, it seems difficult to predict when these
impacts might be large.

The second impact, changing variable scores, is partially addressed in
\cite{DBLP:conf/sat/MohleB19} who utilize the ideas of
\cite{DBLP:journals/jsat/TakRH11} to backtrack to a level where the
decisions would be unchanged. However, if the length of the backtrack
is greater than $100$ there could still be a divergence between the
variable decisions generated by standard backtracking and \cbt. An
argument is also given in \cite{DBLP:journals/jsat/TakRH11} that the
third impact, changing literal reasons, is not significant. However,
the experiments in \cite{DBLP:journals/jsat/TakRH11} were run before
good notions of clause quality were known
\cite{DBLP:conf/ijcai/AudemardS09}. Our empirical results contradict
this, indicating that the literal reasons can have a significant
impact.

The first impact is worth discussing since it was mentioned in the
original \cbt paper \cite{DBLP:conf/lpar/JiangZ13} but not in the
subsequent works. This is the issue of changing the decision levels of
literals on the trail. \cbt computes the decision level of each
implied literal based on the decision levels of its reasons, but it
does not go backwards to change the decisions levels of literals
earlier on the trail.

\begin{example}
    For example, say that $(x, \lnot y)\in \formula$, the literal $x$
    is a decision literal on the trail with $\dlevel{x}=2$, and that
    the solver is currently at level $150$ where it encounters a
    conflict. If this conflict yields the unit clause $(y)$, standard
    backtracking would backtrack to level $0$, where $x$ would be
    implied. On redescent, $x$ would no longer form a new decision
    level and it would not appear in any new clauses (as it is
    entailed by $\formula$). \cbt, on the other hand, would backtrack
    to level $149$. On its trail $x$ would still be at level
    $2$. Until a backtrack past level $2$ occurs, learnt clauses might
    contain $\lnot x$, and thus have level $2$ added to their set of
    levels (potentially changing their LBD score). Only when backtrack
    past level $2$ occurs would $x$ be restored to its correct level
    $0$, and it would require inprocessing simplications to remove $x$
    from the learnt clauses.
\end{example}

In sum, although these impacts of \cbt might or might not be harmful
to the \sat solver, they exist. Hence, one unappealing aspect of \cbt
is that introduces additional inter-dependencies between components of
the solver. For example, the second impact means that heuristics for
choosing decision variables might work differently in the presence of
\cbt; and the first impact means that different ways of computing
clause quality using decision levels in the clauses might also work
differently in the presence of \cbt. So although there are bound to be
cases where \cbt is superior, we would like to have a method that
addresses the original motivation of reducing redundant work while not
changing the operation of standard backtracking. We present our
proposal for such a method next.

\section{Trail Saving}
Our approach is to save the trail, $\trail$, on backtrack, and to use
the saved trail $\trailsave$ when the solver redesends to improve the
effiency of propagations without affecting the decisions the solver
wants to make. The saved trail $\trailsave$ also provides a secondary
``lookahead mechanism'' that the \sat solver can exploit as it
redescends.

Say that the solver is at $\deepestLvl$ where it has encountered a
conflict. From the 1-UIP clause it learns,, $\uipcls$, it now has to
backtrack to $\btL$. This is accomplished by calling the following
subroutine with argument $\btL$.
\begin{algorithmic}[1]
    \Procedure{backtrack}{$\btL$}
    \State $\forall \ell\in\trail[[\range{\btL{+}1}{\deepestLvl}]]$ unassign($\ell$)
    \State $\forall \ell\in\trail[[\range{\btL{+}1}{\deepestLvl{-}1}]]$ $\reasonsave{\ell} = \reason{\ell}$
    \State $\trailsave = \trail[[\range{\btL{+}1}{\deepestLvl{-}1}]]$
    \State $\trail = \trail[[\range{0}{\btL}]]$
    \EndProcedure
\end{algorithmic}

Note that we do not save the deepest level of $\trail$. The full
$\trail$ contains a conflict (at its deepest level). Hence the solver
will never reproduce all the same levels, and it would be useless to
save all of them. Note also that in addition to saving the literals in
$\trailsave$ we also save the clause reason of the unit implied
literals in a $\mathit{reason}_{save}$ vector. Finally, we see that
the first literal on $\trailsave$ must be a decision literal: it is
the first literal of $\trail$ at decision level $\btL+1$.

\paragraph{Using $\trailsave$ for more efficient unit propagations.}
After backtrack the solver will add $\deepestLit$ to the end of the
updated $\trail$ with $\reason(\deepestLit)=\uipcls$ and then invoke
unit propagation which will add more literals to
$\trail$. $\trailsave$ is exploited during propagation by calling the
following version of propagation with the argument
$\trailIdx(\deepestLit)$ (i.e., the trail index of the newly added
implicant.

\begin{algorithm}[t]
\begin{algorithmic}[1]
    \Procedure{unitPropagate}{idx}
    \While{idx $\mbox{} < \trail.\mathrm{size}$()}
        \State $c\gets\mbox{}$ \textsc{useSavedTrail}()
        \If{($c\neq \varnothing$)} \textbf{return} $c$ \Comment{Found conflict from $\trailsave$}
        \EndIf
        \State $\ell \gets \trail[idx]$
        \For{each clause $c\in$ watchlist($\lnot \ell$)}
            \If{($c$ is unit implying $x$)}
                 \State $\trail.\mathrm{addToEnd}(x)$
                 \State $\reason{x} \gets c$            
            \ElsIf{($c$ is falsified)}\:
                 \textbf{return} $c$ \Comment{Found conflict from unit prop.}
            \Else\:
                 Update $c$'s watches.
            \EndIf
        \EndFor
        \State $idx$++
    \EndWhile
    \EndProcedure
    \Statex
    \Procedure{useSavedTrail}{\mbox{}}
    \State $idx\gets 0$; $\litsave\gets \trailsave[idx]$
    \State \qquad Stop at decision literal on $\trailsave$ not yet assigned \true in $\trail$
    \While{($\reasonsave{\litsave} \neq \varnothing \lor \litsave \in \trail$)}
          \If{($\lnot \litsave\in\trail$)}\: \textbf{return} $\reasonsave{\litsave}$ 
                \Comment{Conflict: $\litsave$ already \false}
          \ElsIf{($\litsave \not\in\trail$)} \Comment{$\litsave$ unassigned}
              \State $\trail.\mathrm{addToEnd(\litsave)}$
              \State $\reason{\litsave}\gets \reasonsave{\litsave}$
          \EndIf
          \State $idx$++
    \EndWhile
    \EndProcedure
\end{algorithmic}

\caption{Using $\trailsave$ to enhance unit propagation and conflict
  detection}
\end{algorithm}

\note{
  \begin{enumerate}
  \item Use simple backtrack and save to motivate and state the
      invariant. 
  \item Introduce notation for concatenating two trails.
  \item Name and state the invariant in text (e.g. a
      defn)---concenating the cached trail with the current trail
      yields a trail that is always reason valid, but not necessarily
      non-contradictory, non-conflicting, nor propagation complete).
  \item Proof (simple) that the invariant holds for caching on
      backtrack.
  \item On backtrack at least the asserted literal and any new
      propagants at the backtrack level are added.
  \item Prove that the invariant holds (???do we need an invariant or
      simply a property of the augmented trail) for trails with extras
      inbeween.
      $T_{\mathit{old}} + T_{\mathit{newlyAdded}} +
      T_{\mathit{saved}}$ also satisfies the invariant
  \item re-acheving propagation completeness for the augmented trail
      (a) only add one level of the cached trail at a time, and do
      unit prop after (b) only add non-contradicted decision from the
      old trail. Also note that conflicts are valid and can be learnt
      from even if the augmented trail is not propagation complete. 
  \item Finally figure out a way to formalize the extra technique of
      storing on top after repeated backtracking. 
  \item Explain how freedom for the search is maintained by makeing
      decisions then consulting the trail.
  \item 
  \end{enumerate}
}

@@

% \begin{algorithm}
%   	\TitleOfAlgo{propagate()}
% 	\begin{algorithmic}[1]
%         \While{queue not empty}
%         \Statex
%         \State restoreTrailCache()
%         \Statex
%         \State $l := $ next literal on queue
%         \For{each clause $c \in $ watchlist($\neg l$)}
% 		\For{each literal $x \in c$}
%         \If{$val($x$) \neq$ false and x is not already a watcher of c}
%         \State let $x$ be a watcher of $c$ and add $c$ to watchlist($\neg x$)
%         \State break
%         \EndIf
% 		\EndFor
% 		\If{no new watcher was found}
%         \If{other watcher $x_2$ of $c = $ false}
%         \State return c as conflict
%         \ElsIf{other watcher $x_2$ of $c$ is unassigned}
%         \State enqueue $x_2$
%         \EndIf
% 		\EndIf
%         \EndFor
%         \EndWhile
% 	\end{algorithmic}
% \end{algorithm}

% \begin{algorithm}
%   	\TitleOfAlgo{restoreTrailCache()}
% 	\begin{algorithmic}[1]
%         \If{trail saving turned on}
% 		\State{let tc $=$ top of trail cache}
% 		\If{val(top of trail cache) $=$ true}
%         \While{trail cache not empty}
%         \State{move top of trail cache (tc) to next literal}
%         \If{reason(tc) is null}
%         \State{break}
%         \ElsIf{val(tc) $=$ false}
%         \State{return conflict with reason(tc)}
%         \ElsIf{val(tc) $=$ unassigned}
%         \State{enqueue tc with reason reason(tc)}
%         \EndIf
%         \EndWhile
% 		\EndIf
%         \EndIf
% 	\end{algorithmic}
% \end{algorithm}
